\begin{alltt}
{\it ;;;;}
{\it ;;;; {\bf Utilities}}
{\it ;;;;}
(define make-tk-name 
  (lambda (parent)
    (gensym (format #f "~A.v" (if (eq? parent *root*) "" (Id parent))))))

(define split-options
  (lambda (valid-slots initargs)
    (letrec 
        ((separate 
          (lambda (valids args tk-opt other)
            (if (null? args)
                (cons tk-opt other)
                (if (member (car args) valids)
                    (separate valids (cddr args)
                              (list* (car args) (cadr args) tk-opt)
                              other)
                    (separate valids (cddr args)
                              tk-opt 
                              (list* (car args) (cadr args) other)))))))
      (separate valid-slots initargs '() '()))))

{\it ;;;;}
{\it ;;;; {\bf Simple widgets}}
{\it ;;;;}
{\it ;; }
{\it ;; {\tt{}<}Tk-metaclass{\tt{}>} class definition and associated methods}
{\it ;;}
(define-class {\tt{}<}Tk-Metaclass{\tt{}>} ({\tt{}<}class{\tt{}>})
  ((valid-options :accessor Tk-valid-options)))


(define-method initialize ((class {\tt{}<}Tk-Metaclass{\tt{}>}) initargs)
  (next-method)
  {\it ;; Build a list of allowed keywords. These keywords will be passed to}
  {\it ;; the Tk-command at build time}
  (let ((slots        (slot-ref class 'slots))
        (res         '())
        (tk-virtual?  (lambda(s) 
                        (eqv? (get-slot-allocation s) :tk-virtual))))
    (for-each (lambda (s)
                (when (tk-virtual? s)
                  (let ((key (make-keyword (car s))))
                    (set! res (cons key res)))))
              slots)
    {\it ;; Store this list in the new allocated class}
    (set! (Tk-valid-options class) res)))


(define-method compute-get-n-set ((class {\tt{}<}Tk-Metaclass{\tt{}>}) slot)
  (if (eqv? (get-slot-allocation slot) :tk-virtual)
      {\it ;; this is a Tk-virtual slot}
      (let ((opt (make-keyword (car slot))))
        (list (lambda (o)   (list-ref ((Id o) 'configure opt) 4))
              (lambda (o v) ((Id o) 'configure opt v))))
      {\it ;; call super compute-get-n-set}
      (next-method)))

{\it ;;}
{\it ;; Basic virtual classes for widgets: {\tt{}<}Tk-object{\tt{}>}, {\tt{}<}Tk-widget{\tt{}>} and }
{\it ;; {\tt{}<}Tk-simple-widget{\tt{}>}}
{\it ;;}
(define-class {\tt{}<}Tk-object{\tt{}>} ()
  ((Id      :accessor Id)                             {\it ;; Widget Id}
   (parent  :accessor parent :init-keyword :parent))) {\it ;; Parent widget}

(define-class {\tt{}<}Tk-widget{\tt{}>} ({\tt{}<}Tk-object{\tt{}>})
  ())



(define-class {\tt{}<}Tk-simple-widget{\tt{}>} ({\tt{}<}Tk-widget{\tt{}>})
  {\it ;; Each widget has at least the slot bg for its background colour}
  ((bg :accessor bg :init-keyword :bg :allocation :tk-virtual))
  :metaclass {\tt{}<}Tk-Metaclass{\tt{}>})


(define-method initialize ((self {\tt{}<}Tk-simple-widget{\tt{}>}) initargs)
  {\it ;; Use split-options on initargs to separate STklos slots }
  {\it ;; from Tk ones. Set parent to the root window if not specified}
  {\it ;; in initargs}
  (let* ((options (split-options (Tk-valid-options (class-of self))
                                 initargs))
         (parent  (get-keyword :parent (cdr options) *root*)))
    {\it ;; Call the Tk command which creates the widget}
    (set! (Id self) (apply (tk-constructor self)
                           (make-tk-name parent) 
                           (car options)))
    {\it ;; Initialize other slots (i.e. non Tk-virtual ones)}
    (next-method self (cdr options))))


{\it ;;}
{\it ;; We can now define three widget classes: {\tt{}<}Label{\tt{}>}, {\tt{}<}Button{\tt{}>} and {\tt{}<}Canvas{\tt{}>}}
{\it ;; as well as their associated Tk-command}
{\it ;;}
(define-class {\tt{}<}Label{\tt{}>} ({\tt{}<}Tk-simple-widget{\tt{}>})
  ((font :accessor font :init-keyword :font :allocation :tk-virtual)
   (text :accessor text :init-keyword :text :allocation :tk-virtual)))

(define-class {\tt{}<}Button{\tt{}>} ({\tt{}<}Label{\tt{}>})
  ((command :accessor command :init-keyword :command 
            :allocation :tk-virtual)))

(define-class {\tt{}<}Canvas{\tt{}>} ({\tt{}<}Tk-simple-widget{\tt{}>})
  ())

(define-method tk-constructor ((self {\tt{}<}Label{\tt{}>}))  label)
(define-method tk-constructor ((self {\tt{}<}Button{\tt{}>})) button)
(define-method tk-constructor ((self {\tt{}<}Canvas{\tt{}>})) canvas)


{\it ;;;;}
{\it ;;;; {\bf Canvas items widgets}}
{\it ;;;;}
{\it ;; }
{\it ;; {\tt{}<}Tk-item-metaclass{\tt{}>} class definition and associated methods  }
{\it ;; }

(define-class {\tt{}<}Tk-item-metaclass{\tt{}>} ({\tt{}<}Tk-Metaclass{\tt{}>})
  ())

(define-method compute-get-n-set ((class {\tt{}<}Tk-item-metaclass{\tt{}>}) slot)
  (if (eqv? (get-slot-allocation slot) :tk-virtual)
      {\it ;; this is a Tk-virtual slot}
      (let ((opt (make-keyword (car slot))))
        (list (lambda (obj)   
                (list-ref ((Id obj) 'itemconfigure (Cid obj) opt) 4))
              (lambda (obj val) 
                ((Id obj) 'itemconfigure (Cid obj) opt val))))
      {\it ;; call super compute-get-n-set}
      (next-method)))

{\it ;;}
{\it ;; Basic virtual class: {\tt{}<}Tk-canvas-item{\tt{}>} }
{\it ;;}
(define-class {\tt{}<}Tk-canvas-item{\tt{}>} ({\tt{}<}Tk-object{\tt{}>})
  ((Cid :accessor  Cid)
   (width :accessor width :allocation :tk-virtual))
  :metaclass {\tt{}<}Tk-item-metaclass{\tt{}>})


(define-method initialize ((self {\tt{}<}Tk-canvas-item{\tt{}>}) initargs)
  (let* ((options (split-options (Tk-valid-options (class-of self))
                                 initargs))
         (parent  (get-keyword :parent (cdr options) #f))
         (coords  (get-keyword :coords (cdr options) #f)))
    (if (not (and parent coords))
        (error "Parent widget and coordinates must be given!!"))
    (set! (Id  self) (Id parent))
    (set! (CId self) (apply (Id parent) 
                            'create
                            (canvas-item-initializer self)
                            (append coords (car options))))
    {\it ;; Initialize other slots (i.e. non Tk-virtual ones)}
    (next-method self (cdr options))))

{\it ;;}
{\it ;; We can now define two canvas item classes: {\tt{}<}Line{\tt{}>} and {\tt{}<}Rectangle{\tt{}>}}
{\it ;; as well as their associated initializer}
{\it ;;}
(define-class {\tt{}<}Line{\tt{}>}      ({\tt{}<}Tk-canvas-item{\tt{}>})
  ())

(define-class {\tt{}<}Rectangle{\tt{}>} ({\tt{}<}Tk-canvas-item{\tt{}>})
  ((fill  :accessor fill :init-keyword :fill :allocation :tk-virtual)))

(define-method canvas-item-initializer ((self {\tt{}<}Rectangle{\tt{}>})) "rectangle")
(define-method canvas-item-initializer ((self {\tt{}<}Line{\tt{}>}))      "line")
\end{alltt}
